Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-09-12 foreground_delay_spectrum_modeling.tex
Figure should be capitalized

One important feature of the MWA's measured spectrum is the shape of its bandpass. The band is divided coarsely into 24 chunks with each chunk weighted by a digital filter. The channels at the edges of these chunks have been flagged because they are known to possibly be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in figure~\ref{fig:bandpass}.

@@

We show results of delay spectrum from MWA data after deconvolution along delay axis in figure~\ref{fig:fhd_data}. The principle feature is the wedge--shaped structure, the boundaries of which, marked by solid white lines, increases linearly with light travel time corresponding to the baseline length. Note that the wedge is not filled out uniformly. Each baseline has a different orientation and resulting response to different objects in the sky. It is this variation that we are investigating.

@@

A low resolution version of the diffuse foreground model used is shown in figure~\ref{fig:DSM}. Contours of the MWA tile power--pattern shown in figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the visible hemisphere in the westward sky and the MWA tile's power receptivity is significantly high ($\gtrsim 12$\%) in that direction. In the {\it zenith} pointing, the Galactic plane has almost set and the contour level of power--pattern in that direction is at least 16 times smaller.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
\caption{Sky brightness temperature (in K) of the diffuse foreground model at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power--pattern shown in figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power receptivity is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
\caption{Flux densities (in Jy) of bright compact objects at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power--pattern shown in figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. The locations of bright compact objects appear to be random.\label{fig:CSM}}
\end{figure}

@@

With the aforementioned all--sky foreground model, and instrumental and observational parameters, we simulate visibilities using equation~\ref{eqn:obsvis}. Figure~\ref{fig:sim_data} shows the amplitude of delay spectrum from {\it off--zenith} and {\it zenith} pointings. Notice the qualitative agreement of amplitude and structure with those obtained from data shown in figure~\ref{fig:fhd_data}. The Galactic center and the Galactic plane visible in the {\it off--zenith} pointing make it appear brighter in the {\it foreground wedge}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_sim_delay_spectrum_snapshots.eps}
\caption{Delay spectrum amplitude, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz), obtained with simulations for {\it off--zenith} (top) and {\it zenith} (bottom) pointings. White lines mark the boundaries of {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The features resemble those obtained with MWA data shown in figure~\ref{fig:fhd_data}. The Galactic center and Galactic plane prominently visible in the {\it off--zenith} pointing makes the emission in the {\it foreground wedge} brighter relative to that in the {\it zenith} pointing. The axes and color scale are identical to those in figure~\ref{fig:fhd_data}. \label{fig:sim_data}}
\end{figure}

@@

These uncertainties are presented only to confirm the qualitative agreement already seen between figures~\ref{fig:fhd_data} and \ref{fig:sim_data} and are not intended to serve as a comprehensive estimate of all uncertainties involved. However, since we have not accounted for numerous other uncertainties, especially in the data, such as uncertainties in the antenna power--pattern, calibration, radio frequency interference (RFI), and wide field imaging effects, our estimates are conservative. Accounting for such uncertainties will only improve our confidence level in this comparison. Our primary objective is to explore in detail the foreground signatures embedded in the {\it foreground wedge}.

@@

\item {\it Sky Model}: Our model of the sky consists of bright compact objects and diffuse emission on diverse spatial scales as seen from figures~\ref{fig:DSM} and \ref{fig:CSM}. It also consists of localized regions of strong emission such as the Galactic plane and Galactic center. The resulting sky emission is anisotropic. In fact, the patch of sky for MWA observations is chosen from regions of low foreground emission. In our detailed study, we have divided the transiting sky into four different ``bow-tie'' shaped regions with equal areas around zenith as the origin. This results in the ability to attribute the features in the delay spectrum to eight different sectors of the sky. %TODO possibly reference sections or figures here?

@@

\item {\it Tile Pointing and Power--Pattern}: Since the MWA tiles are steered electronically, the power--pattern of the tile changes in any observing mode that tracks the object. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary. In our study, we take into account the effect of changing power--pattern of the antennas on the delay spectrum (Figure~\ref{fig:power_pattern}). The LST of our data ranges from $\sim$~21 hours through $\sim$~2 hours, and for our analysis we chose {\it off--zenith} and {\it zenith} pointings shown in figure~\ref{fig:pointings}.

@@

The most prominent signature seen in the {\it off--zenith} pointing (top panel of figure~\ref{fig:noiseless-dsm-delay-spectrum}) is due to the bright Galactic center situated on the far west on one of the sidelobes of the power--pattern. It appears as a bright branch in the delay spectrum near the negative delay horizon delay limit. This feature is strongest at short antenna spacings and fades with increase in antenna spacing. This Galactic center signature is almost invisible in the {\it zenith} pointing because the Galactic center has already set over the horizon (see bottom panel in figure~\ref{fig:noiseless-dsm-delay-spectrum}).

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing; each is a different vertical slice of the two dimensional power spectrum plots shown in figures~\ref{fig:fhd_data}, \ref{fig:sim_data}, \ref{fig:noiseless-dsm-delay-spectrum} and \ref{fig:noiseless-csm-delay-spectrum}. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as a pair of vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. Objects inside the {\it foreground} wedge (between the horizon delay limits) are dominated by the $\gtrsim$~70\% error owing to uncertainties in predicting the spectral index of compact foreground objects, outside the wedge thermal noise dominates.

@@

With a foreground model known {\it a priori} in which locations of very bright foreground objects such as the Galactic center or AGN are available, we can predict their response across antenna spacings as a function of observing parameters such as LST, power--pattern, etc. Specifically, we can pick antenna pairs where these bright foreground objects will be visible at delays near horizon delay limits. One such example is already provided in our study. The Galactic center is the most dominant source of foreground contamination in the {\it off--zenith} pointing. Since it is located in the western sky, it is most distinctly seen near the negative horizon delay limit on eastward oriented antenna spacings in the delay spectrum (see Figures~\ref{fig:fhd_data}, \ref{fig:sim_data} and \ref{fig:noiseless-dsm-delay-spectrum}, and \S\ref{sec:GC-east}). In fact these figures illustrate quite well the leakage beyond the {\it foreground wedge} into the {\it EoR window} due to these extremely bright objects near the boundaries of the wedge. In practice, we could downweight these antenna spacings for the specific LST to mitigate the effects of spillover from such foreground contamination in EoR studies. With spatial scales of such objects known, besides downweighting based on antenna spacing orientation, we could also downweight based on antenna spacing distance. For instance, a bright compact object will yield uniform {\it visibility} amplitudes across antenna spacings whereas that of an extended object will gradually fade with increase in antenna spacing.

>>>>

One important feature of the MWA's measured spectrum is the shape of its bandpass. The band is divided coarsely into 24 chunks with each chunk weighted by a digital filter. The channels at the edges of these chunks have been flagged because they are known to possibly be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}. The bandpass properties used in this paper are illustrated in Figure~\ref{fig:bandpass}.

@@

We show results of delay spectrum from MWA data after deconvolution along delay axis in Figure~\ref{fig:fhd_data}. The principle feature is the wedge--shaped structure, the boundaries of which, marked by solid white lines, increases linearly with light travel time corresponding to the baseline length. Note that the wedge is not filled out uniformly. Each baseline has a different orientation and resulting response to different objects in the sky. It is this variation that we are investigating.

@@

A low resolution version of the diffuse foreground model used is shown in Figure~\ref{fig:DSM}. Contours of the MWA tile power--pattern shown in Figure~\ref{fig:power_pattern} are also overlaid. Of notable significance in the {\it off--zenith} pointing is the presence of a portion of the Galactic plane and the bright Galactic center in the visible hemisphere in the westward sky and the MWA tile's power receptivity is significantly high ($\gtrsim 12$\%) in that direction. In the {\it zenith} pointing, the Galactic plane has almost set and the contour level of power--pattern in that direction is at least 16 times smaller.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/dsm.eps}
\caption{Sky brightness temperature (in K) of the diffuse foreground model at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.00195, 0.00781, 0.0312, 0.125, and 0.5. The Galactic center and a portion of the Galactic plane are prominently visible during the {\it off--zenith} pointing and the MWA tile power receptivity is significant ($\gtrsim 12$\%) in that direction. In contrast, emission from the Galactic plane in {\it zenith} pointing is significantly lesser. \label{fig:DSM}}
\end{figure}

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/csm.eps}
\caption{Flux densities (in Jy) of bright compact objects at 185~MHz in Right Ascension ($\alpha$) and Declination ($\delta$) coordinates visible during {\it off--zenith} (top) and {\it zenith} (bottom) pointings. The color scale on the right is logarithmic and is common to both panels and corresponds to the color--coded filled circles. Contours of power--pattern shown in Figure~\ref{fig:power_pattern} are overlaid. The contour levels shown are 0.001953125, 0.0078125, 0.03125, 0.125, and 0.5. The locations of bright compact objects appear to be random.\label{fig:CSM}}
\end{figure}

@@

With the aforementioned all--sky foreground model, and instrumental and observational parameters, we simulate visibilities using equation~\ref{eqn:obsvis}. Figure~\ref{fig:sim_data} shows the amplitude of delay spectrum from {\it off--zenith} and {\it zenith} pointings. Notice the qualitative agreement of amplitude and structure with those obtained from data shown in Figure~\ref{fig:fhd_data}. The Galactic center and the Galactic plane visible in the {\it off--zenith} pointing make it appear brighter in the {\it foreground wedge}.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_sim_delay_spectrum_snapshots.eps}
\caption{Delay spectrum amplitude, $|V_\eta(|\overline{\mathbf{x}}|,\eta)|$ (in units of Jy~Hz), obtained with simulations for {\it off--zenith} (top) and {\it zenith} (bottom) pointings. White lines mark the boundaries of {\it foreground wedge} determined by the horizon delay limit and antenna spacing. The features resemble those obtained with MWA data shown in Figure~\ref{fig:fhd_data}. The Galactic center and Galactic plane prominently visible in the {\it off--zenith} pointing makes the emission in the {\it foreground wedge} brighter relative to that in the {\it zenith} pointing. The axes and color scale are identical to those in Figure~\ref{fig:fhd_data}. \label{fig:sim_data}}
\end{figure}

@@

These uncertainties are presented only to confirm the qualitative agreement already seen between Figures~\ref{fig:fhd_data} and \ref{fig:sim_data} and are not intended to serve as a comprehensive estimate of all uncertainties involved. However, since we have not accounted for numerous other uncertainties, especially in the data, such as uncertainties in the antenna power--pattern, calibration, radio frequency interference (RFI), and wide field imaging effects, our estimates are conservative. Accounting for such uncertainties will only improve our confidence level in this comparison. Our primary objective is to explore in detail the foreground signatures embedded in the {\it foreground wedge}.

@@

\item {\it Sky Model}: Our model of the sky consists of bright compact objects and diffuse emission on diverse spatial scales as seen from Figures~\ref{fig:DSM} and \ref{fig:CSM}. It also consists of localized regions of strong emission such as the Galactic plane and Galactic center. The resulting sky emission is anisotropic. In fact, the patch of sky for MWA observations is chosen from regions of low foreground emission. In our detailed study, we have divided the transiting sky into four different ``bow-tie'' shaped regions with equal areas around zenith as the origin. This results in the ability to attribute the features in the delay spectrum to eight different sectors of the sky. %TODO possibly reference sections or figures here?

@@

\item {\it Tile Pointing and Power--Pattern}: Since the MWA tiles are steered electronically, the power--pattern of the tile changes in any observing mode that tracks the object. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary. In our study, we take into account the effect of changing power--pattern of the antennas on the delay spectrum (Figure~\ref{fig:power_pattern}). The LST of our data ranges from $\sim$~21 hours through $\sim$~2 hours, and for our analysis we chose {\it off--zenith} and {\it zenith} pointings shown in Figure~\ref{fig:pointings}.

@@

The most prominent signature seen in the {\it off--zenith} pointing (top panel of Figure~\ref{fig:noiseless-dsm-delay-spectrum}) is due to the bright Galactic center situated on the far west on one of the sidelobes of the power--pattern. It appears as a bright branch in the delay spectrum near the negative delay horizon delay limit. This feature is strongest at short antenna spacings and fades with increase in antenna spacing. This Galactic center signature is almost invisible in the {\it zenith} pointing because the Galactic center has already set over the horizon (see bottom panel in Figure~\ref{fig:noiseless-dsm-delay-spectrum}).

@@

Figure~\ref{fig:pitchfork-baselines} shows delay spectra of three antenna pairs of different antenna spacings oriented northward during the {\it zenith} pointing; each is a different vertical slice of the two dimensional power spectrum plots shown in Figures~\ref{fig:fhd_data}, \ref{fig:sim_data}, \ref{fig:noiseless-dsm-delay-spectrum} and \ref{fig:noiseless-csm-delay-spectrum}. The diffuse, compact, and composite components are are shown as dotted, dashed, and solid lines respectively. The horizon delay limits are shown as a pair of vertical dot--dashed lines. The gray shaded area denotes the envelope of expected uncertainty in the delay spectrum. Objects inside the {\it foreground} wedge (between the horizon delay limits) are dominated by the $\gtrsim$~70\% error owing to uncertainties in predicting the spectral index of compact foreground objects, outside the wedge thermal noise dominates.

@@

With a foreground model known {\it a priori} in which locations of very bright foreground objects such as the Galactic center or AGN are available, we can predict their response across antenna spacings as a function of observing parameters such as LST, power--pattern, etc. Specifically, we can pick antenna pairs where these bright foreground objects will be visible at delays near horizon delay limits. One such example is already provided in our study. The Galactic center is the most dominant source of foreground contamination in the {\it off--zenith} pointing. Since it is located in the western sky, it is most distinctly seen near the negative horizon delay limit on eastward oriented antenna spacings in the delay spectrum (see Figures~\ref{fig:fhd_data}, \ref{fig:sim_data} and \ref{fig:noiseless-dsm-delay-spectrum}, and \S\ref{sec:GC-east}). In fact these Figures illustrate quite well the leakage beyond the {\it foreground wedge} into the {\it EoR window} due to these extremely bright objects near the boundaries of the wedge. In practice, we could downweight these antenna spacings for the specific LST to mitigate the effects of spillover from such foreground contamination in EoR studies. With spatial scales of such objects known, besides downweighting based on antenna spacing orientation, we could also downweight based on antenna spacing distance. For instance, a bright compact object will yield uniform {\it visibility} amplitudes across antenna spacings whereas that of an extended object will gradually fade with increase in antenna spacing.
